Well-quasi-orderings on word languages

### 2024-08-06 src/infixes.tex
rewriting introduction

In this section, we study the well-quasi-ordering of languages under the infix
relation. As opposed to the prefix and suffix relations, the infix relation can
yield to very complicated well-quasi-ordered languages. As a matter of fact,
the following theorem shows that \emph{any} countable quasi-ordering with
finite initial segments can be embedded into the infix relation of a language.

>>>>

In this section, we study the well-quasi-ordering of languages under the
\kl{infix relation}. As opposed to the \kl{prefix} and \kl{suffix} relations,
the \kl{infix relation} can yield to very complicated \kl{well-quasi-ordered}
languages. Formally, the following theorem shows that \emph{any} countable
quasi-ordering with finite initial segments can be embedded into the infix
relation of a language.

### 2024-08-06 src/introduction.tex
rewriting introduction

\section{Introduction}
\label{introduction:sec}

\subparagraph{Motivation}

>>>>

% LTeX: language=en-GB 
\section{Introduction}
\label{introduction:sec}

A \intro{well-quasi-ordered} set is a set $X$ equipped with a quasi-order
$\preceq$ such that every infinite sequence $\seqof{x_n}$ of elements taken in
$X$ contains a pair $x_i \preceq x_j$ with $i < j$. Well-quasi-orderings serve
as a core combinatorial tool powering many termination arguments, and was
successfully applied to the verification of infinite state transition systems
\cite{ABDU96,ABDU98}. One of the appealing properties of well-quasi-orderings
is that they are closed under many operations, such as taking products, finite
unions, and finite powersets \cite{SCSC12}. More surprisingly, the class of
well-quasi-ordered sets is also stable under the operation of taking finite
words and finite trees labelled by elements of a well-quasi-ordered set
\cite{HIG52,KRU72}.

Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. The three alternatives
we consider are the \kl{prefix relation}, the \kl{suffix relation}, and the
\kl{infix relation}. Unfortunately, it is an easy check that all of these
constructions are not well-quasi-ordered as soon as the alphabet contains two
non-equivalent letters. This is not the end of the road, because some
\emph{subsets} of $\Sigma^*$ might still be \kl{well-quasi-ordered} by these
relations, the simplest example being the case of finite sets of (finite) words
which are all \kl{well-quasi-ordered} regardless of the ordering considered.
This raises the question of characterizing the subsets $L \subseteq \Sigma^*$
that are \kl{well-quasi-ordered} with respect to the \kl{prefix relation}
(respectively, the \kl{suffix relation} or the \kl{infix relation}), and
providing suitable decision procedures.

\subparagraph{Contribution} In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Let us remark that these word orderings are all
\intro{well-founded}, i.e., they do not contain infinite \intro{decreasing
sequences} $\seqof{x_n}$ such that $x_{n+1} \prec x_n$ for all $n \in \Nat$.
Let us recall that the key difference between \kl{well-quasi-orderings} and the
more usual notion of well-founded orderings is that the latter allows the
existence of infinite \intro{antichains}, that is, sets of pairwise
incomparable elements. In fact, a set $X$ is \kl{well-quasi-ordered} if and
only if it is \kl{well-founded} and does not contain an infinite \kl{antichain}
\cite{KRU72}.  As a consequence, our study focuses on detecting infinite
\kl{antichains} in languages $L \subseteq \Sigma^*$.

In the case of the \kl{prefix relation}, we provide a complete characterization
of languages $L \subseteq \Sigma^*$ that are \kl{well-quasi-ordered}, which are
the simplest possible languages: they are precisely the \emph{finite} unions of
\intro{chains}, that is, sets of words that are all comparable for the
\kl{prefix relation}. Note that \kl{chains} are the simplest possible
\kl{well-quasi-ordered} sets (they are totally ordered and well-founded) and it
is known that finite unions of \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered}. As a consequence, the above characterization states
that only the simplest possible \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered} by the \kl{prefix relation}. Let us furthermore
highlight that this characterization holds without any restriction on the
decidability of the language $L$ itself, but heavily relies on the assumption
that $\Sigma$ is finite. By symmetry, we obtain an analogue statement for the
\kl{suffix relation}.

In the case of the \kl{infix relation}, the situation is more complex, and we
show that \emph{any} countable quasi-ordering with finite initial segments can
be embedded into the \kl{infix relation} of a language over a binary alphabet
$\Sigma \defined \set{a,b}$. As an example, on can embed $(\Sigma^*, \higleq)$
into $(\Sigma^*, \infleq)$ where $\higleq$ is the \kl{subword relation} and
$\infleq$ is the \kl{infix relation}. This strongly contrasts with the
\kl{prefix relation}, and suggests that a meaningful characterization of
\kl{well-quasi-ordered} languages under the \kl{infix relation} is unlikely to
be found.

We circumvent this issue by adding some regularity assumptions on the languages
$L \subseteq \Sigma^*$ that we consider. This can be done by assuming that $L$
is recognized by a finite automaton, or more generally that $L$ is the language
of an \kl{amalgamation system}. Under these relatively mild assumptions, we
also prove that \kl{well-quasi-ordered} languages under the \kl{infix relation}
are precisely the finite unions of \kl{chains}.

\subparagraph{Related work} The study of alternative \kl{well-quasi-ordered}
relations over finite words is far from new. For instance, orders obtained by
so-called \emph{derivation relations} where already analysed by Bucher,
Ehrenfeucht, and Haussler \cite{BUEUD85}. These results were later extended by
D'Alessandro and Varricchio \cite{ALVA03,ALVA06}. However, in all those cases
the orderings are \emph{multiplicative}, that is, if $u_1 \preceq v_1$ and $u_2
\preceq v_2$ then $u_1u_2 \preceq v_1v_2$. This assumption does not hold for
the \kl{prefix}, \kl{suffix}, and \kl{infix} relations.

\subparagraph{Outline} In \cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} prove in \cref{prefixes:thm}
the characterization of \kl{well-quasi-ordered} languages by the \kl{prefix
relation}. In \cref{infixes:sec}, we prove that any countable quasi-ordering
with finite initial segments can be embedded into the \kl{infix relation} of a
language (\cref{count-embed-infix:thm}), we obtain the analogue of
\cref{prefixes:thm} for languages recognized by finite automata
(\cref{infix-finite-automata:thm}). Finally, in \cref{amalgamation:sec}, we
introduce \kl{amalgamation systems} and generalize
\cref{infix-finite-automata:thm} to this setting
(\cref{infix-amalgamation:thm}).

\subparagraph{Acknowledgements} We would like to thank participants of the 2024
edition of Autobóz for their helpful comments and discussions.

### 2024-08-06 src/prefixes.tex
rewriting introduction

In this section, we study the well-quasi-ordering of languages under the prefix
relation. A word $u$ is a \intro{prefix} of a word $w$ if there exists a word
$v$ such that $w = uv$. We denote this relation by $u \intro*\prefleq w$.
Similarly, a word $u$ is a \intro{suffix} of a word $w$ if there exists a word
$v$ such that $w = vu$. We denote this relation by $u \intro*\suffleq w$.

Let us immediately remark that the map $u \mapsto u^R$ that reverses a word is
an order-bijection between $(X^*, \prefleq)$ and $(X^*, \suffleq)$. Therefore,
we will focus on the prefix relation in the following.

As mentioned in the introduction, the prefix relation is not a well-quasi-order
on words over a two-letter alphabet, as witnessed by the infinite
\kl{antichain} $\setof{a^nb}{n \in \Nat}$.

Let us order the words in $\Sigma^*$ by the \kl{prefix relation} to build an
infinite tree $T$. The root of the tree is the empty word $\varepsilon$, and
the children of a word $w$ are the words $wa$ for all $a \in \Sigma$. This tree
is finitely branching. Let $L \subseteq \Sigma^*$ be a language. Assume that
$L$ contains an infinite \kl{antichain} $A$ for the \kl{prefix relation}. In
particular, no two word of $A$ belong to the same branch of the tree $T$. Let
us prove the following: there exists an infinite branch $B$ of $T$ such that from
every point of the branch, one can reach a word of $A$ that is not in the
branch.

@@

@@

@@

\begin{corollary}
    Let $L \subseteq \Sigma^*$ be a language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists a finite $k \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $k$.
\end{corollary}

>>>>

In this section, we study the well-quasi-ordering of languages under the prefix
relation. A word $u$ is a \intro{prefix} of a word $w$ if there exists a word
$v$ such that $w = uv$. We denote this relation by $u \intro*\prefleq w$.
Similarly, a word $u$ is a \intro{suffix} of a word $w$ if there exists a word
$v$ such that $w = vu$. We denote this relation by $u \intro*\suffleq w$. Let
us immediately remark that the map $u \mapsto u^R$ that reverses a word is an
order-bijection between $(X^*, \prefleq)$ and $(X^*, \suffleq)$. Therefore, we
will focus on the prefix relation in the rest of this section.

Let us briefly restate the fact that some (even regular) languages 
are not well-quasi-ordered by the prefix relation.

\begin{example}
    The set $L = \setof{a^nb}{n \in \Nat}$ is an infinite antichain for the
    prefix relation.
\end{example}

In order to characterize the existence of infinite antichains for the prefix
relation, we will introduce the following tree construction that
will be useful in the rest of this section.

\begin{definition}
    The \intro{tree of prefixes} over a finite alphabet $\Sigma$
    is the infinite tree $T$ whose nodes are the words of $\Sigma^*$, and
    such that the children of a word $w$ are the words $wa$ for all $a \in
    \Sigma$. 
\end{definition}

Notice that the tree of prefixes is finitely branching. Let us now
observe how antichains in the prefix relation can be witnessed
by infinite branches in the tree of prefixes.

@@

It is clear that an \kl{antichain branch} for a language $L$ yields an infinite
antichain, and the converse is quite easy to prove. Because the notion of
antichain branch is \kl{$\MSO$-definable} whenever $L$ is regular, we
immediately obtain decidability as a corollary of this lemma.

@@

Let us now go further and fully characterize languages $L$ such that the
prefix relation is well-quasi-ordered, even without any restriction on the
decidability of $L$ itself. Let us remark that finite unions of \kl{chains} are
always \kl{well-quasi-ordered} by the \kl{prefix relation} because they lack
infinite \kl{antichains} by definition. The following theorem states that this
is the only possible reason for a language $L$ to be \kl{well-quasi-ordered} by
the \kl{prefix relation}.

@@

As an immediate consequence, we have a very fine-grained understanding of the
\kl{ordinal invariants} of such \kl{well-quasi-ordered} languages, which can be
leveraged in bounding the complexity of algorithms working on such languages.

\begin{corollary}
    \textbf{false: some chains can be finite}
    Let $L \subseteq \Sigma^*$ be a language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists a finite $k \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $k$.
\end{corollary}

Let us conclude by noting that it is unsurprisingly not possible to decide
whether a decidable language is \kl{well-quasi-ordered} by the \kl{prefix
relation}.

\begin{lemma}
    todo.
\end{lemma}
